---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cori.charts

<!-- badges: start -->
<!-- badges: end -->

The goal of cori.charts is to make the process of creating publication-ready graphics and charts at the Center for Rural Innovation as easy as possible. Here’s how to start.

## Installation

You can install the development version of cori.charts from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ruralinnovation/cori.charts")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(cori.charts)
cori_colors["Emerald"]
```


## Contributing
  
Please note that the cori.charts project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
